---
title: "Data Investigation"
output: html_document
---


```{r}
library(tidyverse)
```

## Problem Statement
DonorsChoose.org has funded over 1.1 million classroom requests through the support of 3 million donors, the majority of whom were making their first-ever donation to a public school. If DonorsChoose.org can motivate even a fraction of those donors to make another donation, that could have a huge impact on the number of classroom requests fulfilled.

A good solution will enable DonorsChoose.org to build targeted email campaigns recommending specific classroom requests to prior donors. Part of the challenge is to assess the needs of the organization, uncover insights from the data available, and build the right solution for this problem. Submissions will be evaluated on the following criteria:

Performance - How well does the solution match donors to project requests to which they would be motivated to donate? DonorsChoose.org will not be able to live test every submission, so a strong entry will clearly articulate why it will be effective at motivating repeat donations.

Adaptable - The DonorsChoose.org team wants to put the winning submissions to work, quickly. Therefore a good entry will be easy to implement in production.

Intelligible - A good entry should be easily understood by the DonorsChoose.org team should it need to be updated in the future to accommodate a changing marketplace.

## Website investigation

* Search for topics, teachers and schools near a particular location
* Certain projects are highlighted as urgent. 
* Can see amount left to fund projects, number of donors, goal amount
* Expires date is there but not obvious
* Has a profile of teacher with name and picture
* Projects only go through if they become fully funded
* Can see dollar break down of where your money would go. 
* Goal amount includes the suggested donation to help donors choose reach more classrooms
* Ripple sometimes matches donations
* Highlights when donations are matched

## Data Loading

Load data from CSV (over 3 GBs of files)
```{r}
donations <- read_csv("data/Donations.csv")
donors <- read_csv("data/Donors.csv")
projects <- read_csv("data/Projects.csv")
resources <- read_csv("data/Resources.csv")
schools <- read_csv("data/Schools.csv")
teachers <- read_csv("data/Teachers.csv")
```

## Data investigation
Look at the data to see what the values are and what is missing.
#### Donations
Look into donations: 
Has a bunch of FKs, "Donation Included Optional Donation" which should be a logical, donation amount (in usd?), "Donor Cart Sequence" which is the nth checked out cart for that donor
```{r}
head(donations, n = 25)
```

Most donations include optional donation (On their website it indicates that this is an optional extra donation which people can make, to support the operation of DonorsChoose, rather than the project itself). Amounts can be 0 or NA? Not sure what that means. Most donors are in their first few carts but someone has checked out 18000 times? That seems weird. Very large donations but seems reasonable.
```{r}
summary(donations %>% mutate(`Donation Included Optional Donation` = as.factor(`Donation Included Optional Donation`)))
```

5 donors with more that 5000 carts. Note that these skew the mean cart because every intermediate cart is counted as well.
```{r}
donations %>% filter(Donor.Cart.Sequence > 5000) %>% group_by(Donor.ID) %>% count()
```



#### Donors

Contains basic info on donors location and whether they are a teacher. This would be far more valuable if we could get some data with more demographic data by location. Could see impact of income on donations. Are these only whether they are teachers in their system or if they are teachers at all?
```{r}
head(donors, n=30)
```

More people are not teachers than are. Lots of people missing zip code.
```{r}
summary(donors %>% mutate(`Donor Is Teacher` = as.factor(`Donor Is Teacher`), `Donor Zip` = as.factor(`Donor Zip`)))
```

More people missing city than zip code which seems weird. 
```{r}
sapply(donors, function(x) sum(is.na(x)))
```

Seem to actually have valid zip codes so could probably use data to create a city to zipcode map and fill in missing cities
```{r}
donors %>% filter(is.na(`Donor City`), !is.na(`Donor Zip`)) %>% head()
```

#### Projects
Gives a list of donation campaigns with info about them. "Teacher Project Posted Sequence" indicates the nth project for that teacher. Project type is a category with few bins. Project Essay is where most of the size comes from. Could do some nlp on these. Has more project info like funding status and categories. Seems to have some data cleanliness issues. Project cost $ prefixed so should remove. "Archived" means the project was removed from the site by a staff member or by the teacher. Please ignore these for now- we are updating the data soon to get rid of these projects from the competition dataset.
```{r}
head(projects, n=30)
```

```{r}
summary(projects )
```

Lots missing fully funded data (makes sense). Other missing data may be a csv formatting issue.
```{r}
sapply(projects, function(x) sum(is.na(x)))
```

#### Resources
For every project in the above dataset, there are one or more resources that is requested. Has a text resource name, number, price and vendor. Not sure if these make up the whole cost of a project but most are pretty cheap so probably not? Each project does seem to have a lot of them though.
```{r}
head(resources, n=30)
```

Resource quantity and price sometimes missing. Some prices and quantities are zero which is weird.
```{r}
summary(resources)
```

#### Schools

Lots of info on school like whether they have a high percentage of free lunch students (potential economic indicator), location (do donors donate more / more often when they are closer), 
```{r}
head(schools, n=30)
```

School lunch percentage looks ok, most schools in and around cities.
```{r}
summary(schools %>% mutate(`School Metro Type` = as.factor(`School Metro Type`)))
```

#### Teachers
just gives an optional gender (prefix) and start date. More experienced users might get more funded projects. 
```{r}
head(teachers, n=30)
```

## Data Transformation and Cleaning

Apply some basic data cleaning and formatting

Fix all the column names
```{r}
colnames(donations) <- make.names(colnames(donations))
colnames(donors) <- make.names(colnames(donors))
colnames(projects) <- make.names(colnames(projects))
colnames(resources) <- make.names(colnames(resources))
colnames(schools) <- make.names(colnames(schools))
colnames(teachers) <- make.names(colnames(teachers))
```

Format data correctly
```{r}
donations <- donations %>%
  mutate(Donation.Included.Optional.Donation = Donation.Included.Optional.Donation == 'Yes')

donors <- donors %>%
  mutate(Donor.Is.Teacher = Donor.Is.Teacher == 'Yes',
         Donor.City = as.factor(Donor.City),
         Donor.State = as.factor(Donor.State),
         Donor.Zip = as.factor(Donor.Zip))

projects <- projects %>% 
  mutate(Project.Type = as.factor(Project.Type),
         Project.Subject.Category.Tree = as.factor(Project.Subject.Category.Tree),
         Project.Subject.Subcategory.Tree = as.factor(Project.Subject.Subcategory.Tree),
         Project.Grade.Level.Category = as.factor(Project.Grade.Level.Category),
         Project.Resource.Category = as.factor(Project.Resource.Category),
         Project.Current.Status = as.factor(Project.Current.Status))

resources <- resources %>% 
  mutate(Resource.Vendor.Name = as.factor(Resource.Vendor.Name))

schools <- schools %>% 
  mutate(School.Metro.Type = as.factor(School.Metro.Type),
         School.State = as.factor(School.State),
         School.Zip = as.factor(School.Zip),
         School.City = as.factor(School.City),
         School.County = as.factor(School.County))

teachers <- teachers %>% 
  mutate(Teacher.Prefix = as.factor(Teacher.Prefix))
```

Fill in missing data where applicable
```{r}
donors %>% group_by(Donor.Zip) %>%
  mutate(most_common_city = Donor.City %>% table %>% which.max %>% names)

```

## Data Visualization

### Donations

Very spiky distribution, clearly some amounts like 10, 25, 50, 100 that peopl donate far more than others. Would be useful to bin these. 
```{r}
donations %>% ggplot(aes(x=Donation.Amount)) + geom_density() + scale_x_log10()
```

I hypothesize that people who give multiple times are more willing to give less on subsequent donations. Graph is hard to read but seems to show that.
```{r}
donations %>% ggplot(aes(x=Donation.Amount, fill=Donor.Cart.Sequence > 1)) + geom_density(alpha=0.4) + scale_x_log10()
```

Exponential looking drop off with long tail
```{r}
donations %>% filter(Donor.Cart.Sequence < 100) %>% ggplot(aes(x=Donor.Cart.Sequence)) + geom_histogram(binwidth = 1) 
```

Vast majority of donors only donate once and very few donate more than ten times. 
```{r}
donations %>% filter(Donor.Cart.Sequence < 100) %>%
  group_by(Donor.ID) %>% 
  summarise(Donor.Cart.Sequence = max(Donor.Cart.Sequence)) %>% 
  ggplot(aes(x=Donor.Cart.Sequence)) + geom_histogram(binwidth = 1) 
```

General down trend in median as donor cart increases. Increase at first though.
```{r}
donations %>% filter(Donor.Cart.Sequence < 100) %>%
  ggplot(aes(x=Donor.Cart.Sequence, y=Donation.Amount)) +  stat_summary(fun.y=median, geom="line",lwd=2,aes(group=1))
```

#### Projects

Looks like there are a few most common funding targets. Cheaper projects are more likely to get fully funded. Wonder how much projects that are unfunded get on average. 
```{r}
projects %>% 
  filter(Project.Current.Status != "Live") %>% 
  ggplot(aes(x=Project.Cost, fill=Project.Current.Status)) + geom_density(alpha=0.5) + scale_x_log10()
```

Hard to read labels and should fix that but does appear to be some variation in the funding rate for different categories
```{r}
projects %>% 
  filter(Project.Current.Status != "Live") %>%
  ggplot(aes(x=Project.Resource.Category, fill=Project.Current.Status)) + geom_bar(position = "fill")
  
```

Does not appear to be much variation based on grade level. Unknown is less which makes sense.
```{r}
projects %>% 
  filter(Project.Current.Status != "Live") %>%
  ggplot(aes(x=Project.Grade.Level.Category, fill=Project.Current.Status)) + geom_bar(position = "fill")
```

These seem to need to be split on commas as lots have two. The most common are single subjects.
```{r}
projects %>% group_by(Project.Subject.Category.Tree) %>%  count() %>% arrange(desc(n))

projects %>% 
  filter(Project.Current.Status != "Live") %>%
  ggplot(aes(x=Project.Subject.Category.Tree, fill=Project.Current.Status)) + geom_bar(position = "fill")
```
